Simple Type Theory is not too Simple

### 2021-04-07 article.tex

It is well-known in the Isabelle community that the algebra library should be rebuilt from scratch. First, I should say the algebra libraries. Indeed, there are many libraries for algebra in Isabelle, the \textit{HOL-Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Algebra/index.html}} started in the previous millennium (1999), the more recent \textit{HOL-Computational\_Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Computational_Algebra/index.html}}, the oldest formalization of algebraic structures (groups, rings, modules, fields, vector spaces \dots) being found in the main library \textit{HOL}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL/index.html}}, and this is in addition to some entries in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/} like \textit{Groups, Rings and Modules} \cite{Group-Ring-Module-AFP} and \textit{Vector Spaces} \cite{VectorSpace-AFP} to cite a few. As a result, there is an important amount of duplicated code and it makes the situation unclear regarding the status of algebra for the newcomer in the Isabelle world. Second, if these formalizations constitute an impressive body of work, not without its merits, parts of it are deprecated code not using the Isar language or the modern locales, a powerful module system in Isabelle for managing theory hierarchies \cite{ballarin2014}. Therefore, HOL-Algebra contains some peculiarities, a notable one being its locale \emph{abelian\_group} (in \textit{Theory Ring}) which presents the odd twist of requiring a ring structure. Indeed, this last locale uses records and there does not exist multiple inheritances for records, hence the awkward use of the ring record for abelian groups. However, a successful experiment was conducted recently in \cite{ballarin20} showing that modern locales, without records, allow for a smooth handling of basic algebraic structures in Isabelle. So, hopefully, the situation should dramatically improved in the near future. \\

@@

For the remainder of this text fix a commutative ring $R$. \\
The definition of an ideal of $R$ was introduced in Isabelle in \cite{ballarin20}.

@@

Next, we take a detour into the localization of commutative rings. This construction was already formalized by the first author in \cite{bordg18}, but we give a more concise formalization here since the first author gained a better knowledge of the locale mechanism in the meantime. \\
Let $S$ be a multiplicative submonoid of $R$. We consider pairs $(r, s)$ with $r \in R$ and $s \in S$ and we define the following relation on them
	\[
	(r, s) \sim (r', s')
	\]
if and only if there exists $s_1 \in S$ such that $s_1(s' r - s r') = 0$. One checks that the relation $\sim$ is an equivalence relation. The equivalence class of a pair $(r, s)$ is denoted by $r/s$ and the set of equivalence classes is denoted by $S^{-1} R$. \\
We define the following addition on $S^{-1} R$.
	\[
	\frac{r}{s} + \frac{r'}{s'} = \frac{r s' + r' s}{s s'}
	\]
We also define the following multiplication on $S^{-1} R$. 
	\[
	\frac{r}{s} \times \frac{r'}{s'} = \frac{r r'}{s s'}
	\]
One checks $(S^{-1} R, 0/1, +, 1/1, \times)$ is a ring. The new ring $S^{-1} R$ is called the \emph{quotient ring of $R$ by $S$}. \\
This construction translates into the following very legible locale ending with a proof that the quotient ring is indeed a ring.

@@

For our formalization we built upon Ballarin's work \cite{ballarin20} formalizing parts of Jacobson's \textit{Basic Algebra} textbook. We believe Ballarin's nice formalization could serve as a first stepping stone to a new algebra library for Isabelle. We also hope the present work could serve as a second stepping stone in that direction. Also, we seized the opportunity of formalizing schemes to build a new topology library despite the two existing formalizations of topology in Isabelle/HOL, namely \textit{HOL-Topological\_Spaces.thy}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL/Topological_Spaces.html}} and \textit{HOL-Analysis/Abstract\_Topology}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Analysis/Abstract_Topology.html}}. Our new topology library is entirely built on modern locales without using type classes nor type declarations (via the \textit{typedef} command). In particular, our topological spaces have explicit carrier sets as part of their data instead of using \textit{UNIV} or having to define it later as the union of all the open sets. 

>>>>

It is well-known in the Isabelle community that the algebra library should be rebuilt from scratch. First, I should say the algebra libraries. Indeed, there are many libraries for algebra in Isabelle, the \textit{HOL-Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Algebra/index.html}} started in the previous millennium (1999), the more recent \textit{HOL-Computational\_Algebra}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Computational_Algebra/index.html}}, the oldest formalization of algebraic structures (groups, rings, modules, fields, vector spaces \dots) being found in the main library \textit{HOL}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL/index.html}}, and this is in addition to some entries in the Archive of Formal Proofs\footnote{https://www.isa-afp.org/} like \textit{Groups, Rings and Modules} \cite{Group-Ring-Module-AFP} and \textit{Vector Spaces} \cite{VectorSpace-AFP} to cite a few. As a result, there is an important amount of duplicated code and it makes the situation unclear regarding the status of algebra for the newcomer in the Isabelle world. Second, if these formalizations constitute an impressive body of work, not without its merits, parts of it are deprecated code not using the Isar language or the modern locales, a powerful module system in Isabelle for managing theory hierarchies \cite{ballarin-locales-module}. Therefore, HOL-Algebra contains some peculiarities, a notable one being its locale \emph{abelian\_group} (in \textit{Theory Ring}) which presents the odd twist of requiring a ring structure. Indeed, this last locale uses records and there does not exist multiple inheritances for records, hence the awkward use of the ring record for abelian groups. However, a successful experiment was conducted recently in \cite{ballarin-exploring} showing that modern locales, without records, allow for a smooth handling of basic algebraic structures in Isabelle. So, hopefully, the situation should dramatically improved in the near future. \\

@@

For the remainder of this text fix a commutative ring $R$. \\
The definition of an ideal of $R$ was introduced in Isabelle in \cite{ballarin-exploring}.

@@

Next, we take a detour into the localization of commutative rings. This construction was already formalized by the first author in \cite{Localization_Ring-AFP}, but we give a more concise formalization here since the first author gained a better knowledge of the locale mechanism in the meantime. \\
Let $S$ be a multiplicative submonoid of $R$. We consider pairs $(r, s)$ with $r \in R$ and $s \in S$ and we define the following relation on them
	\[
	(r, s) \sim (r', s')
	\]
if and only if there exists $s_1 \in S$ such that $s_1(s' r - s r') = 0$. One checks that the relation $\sim$ is an equivalence relation. The equivalence class of a pair $(r, s)$ is denoted by $r/s$ and the set of equivalence classes is denoted by $S^{-1} R$. \\
We define the following addition on $S^{-1} R$.
	\[
	\frac{r}{s} + \frac{r'}{s'} = \frac{r s' + r' s}{s s'}
	\]
We also define the following multiplication on $S^{-1} R$. 
	\[
	\frac{r}{s} \times \frac{r'}{s'} = \frac{r r'}{s s'}
	\]
One checks $(S^{-1} R, 0/1, +, 1/1, \times)$ is a ring. The new ring $S^{-1} R$ is called the \emph{quotient ring of $R$ by $S$}. \\
This construction translates into the following very legible locale ending with a proof that the quotient ring is indeed a ring.

@@

For our formalization we built upon Ballarin's work \cite{ballarin-exploring} formalizing parts of Jacobson's \textit{Basic Algebra} textbook. We believe Ballarin's nice formalization could serve as a first stepping stone to a new algebra library for Isabelle. We also hope the present work could serve as a second stepping stone in that direction. Also, we seized the opportunity of formalizing schemes to build a new topology library despite the two existing formalizations of topology in Isabelle/HOL, namely \textit{HOL-Topological\_Spaces.thy}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL/Topological_Spaces.html}} and \textit{HOL-Analysis/Abstract\_Topology}\footnote{\url{https://isabelle.in.tum.de/dist/library/HOL/HOL-Analysis/Abstract_Topology.html}}. Our new topology library is entirely built on modern locales without using type classes nor type declarations (via the \textit{typedef} command). In particular, our topological spaces have explicit carrier sets as part of their data instead of using \textit{UNIV} or having to define it later as the union of all the open sets. 
